On the probability of finding nonphysical solutions through shadowing\tnoteref{1}

### 2020-12-12 main_els.tex

\section{Introduction}
In this paper, we show that a large mismatch is possible in the ensemble averages of two slightly different chaotic systems. We construct numerical and analytical examples that illustrate this feature of chaotic dynamics. The analysis of the long-term behavior of {\em shadowing} solutions in the majority of the chaotic models constructed in this paper reveals that shadowing solutions are almost surely {\em nonphysical}: they do not represent the behavior of a physically observed solution starting almost everywhere. This also leads us to the conclusion that shadowing-based sensitivities \cite{qiqi-lss} of long-time averages to parameter changes can be incorrect. Thus, the main conjecture of this paper urges us to reconsider the use of chaotic numerical simulations, and statistical sensitivity analysis on these simulations. These implications are particularly important for climate studies, where chaotic models are widely used to make long-term predictions. In particular, three implications of the evidence in this paper may be stated, the third of which harnesses the statistical inconsistency between slightly different models:
\begin{itemize}
    \item we cannot trust the long-time averages generated by numerical simulations of chaotic processes, even when we eliminate model uncertainties stemming from imperfect modeling of the underlying physics, and statistical noise due to finite-time averaging;
    \item we cannot trust the sensitivities of long-time averages or statistical averages to small parameter changes computed by using shadowing-based methods for sensitivity computation;
	\item it is possible for two models separated by tiny changes in parameter values to have drastically different statistical behavior. This means that the Lorenzian butterfly (a small perturbation) can significantly alter the climate of Texas (statistical/long-term average), and not just its weather (short-term events).
\end{itemize}
The paper is organized as follows. In section \ref{sec:shadowingIntro}, we introduce the dynamical systems concept of shadowing and its applications in the context of chaotic numerical simulations. In section \ref{sec:nonphysical}, we define the notion of physicality and give examples to illustrate the different ways in which to obtain nonphysical solutions. Section \ref{sec:shadowingnonphysical} is dedicated to several examples of perturbed tent maps that have nonphysical shadowing solutions. In the discussion in section \ref{sec:discussion}, we illustrate with an example that a small parameter perturbation can change the statistics significantly. The supplementary material, section \ref{sec:supp}, contains the numerical procedure to generate the probability distributions of the shadowing solutions.

>>>>

\section{Introduction}
In this paper, we show that a large mismatch is possible in the ensemble averages of two slightly different chaotic systems. We illustrate this feature of chaotic dynamics by studying the long-time behavior of {\em shadowing} solutions in analytically or numerically constructed examples.{\em Shadowing} solutions are true orbits that closely follow a slightly perturbed solution, say due to a small parameter perturbation in the original governing equation, for a long, and sometimes infinitely long, time. In this paper, we show through rigorous analytical and numerical examples that these shadowing orbits can be statistically mismatched. In other words, the probability distribution of shadowing orbits is different, often drastically so, from that of almost every solution.

The analysis of the long-term behavior of {\em shadowing} solutions in the majority of the chaotic models constructed in this paper reveals that shadowing solutions are almost surely {\em nonphysical}: they do not represent the behavior of a physically observed solution starting almost everywhere. This also leads us to the conclusion that shadowing-based sensitivities \cite{qiqi-lss} of long-time averages to parameter changes can be incorrect. Thus, the main conjecture of this paper urges us to reconsider the use of chaotic numerical simulations, and statistical sensitivity analysis on these simulations. These implications are particularly important for climate studies, where chaotic models are widely used to make long-term predictions. In particular, three implications of the evidence in this paper may be stated, the third of which harnesses the statistical inconsistency between slightly different models:

Owing to the {\em shadowing} property of dynamical systems, it is generally believed that a numerical simulation of chaotic governing equation, which are small perturbations of the true solution, are {\em shadowed} or closely followed by a true solution of the physics.

In this paper, we show that a large mismatch is possible in the statistical or long-time averages measured in a pair of chaotic solutions to slightly different models.  that  We construct numerical and analytical examples that illustrate this feature of chaotic dynamics. The analysis of the long-term behavior of {\em shadowing} solutions in the majority of the chaotic models constructed in this paper reveals that shadowing solutions are almost surely {\em nonphysical}: they do not represent the behavior of a physically observed solution starting almost everywhere. This also leads us to the conclusion that shadowing-based sensitivities \cite{qiqi-lss} of long-time averages to parameter changes can be incorrect. Thus, the main conjecture of this paper urges us to reconsider the use of chaotic numerical simulations, and statistical sensitivity analysis on these simulations. These implications are particularly important for climate studies, where chaotic models are widely used to make long-term predictions. In particular, three implications of the evidence in this paper may be stated, the third of which harnesses the statistical inconsistency between slightly different models:
>>>>>>> origin/overleaf-2020-12-12-1530
\begin{itemize}
    \item we cannot trust the long-time averages generated by numerical simulations of chaotic processes, even when we eliminate model uncertainties stemming from imperfect modeling of the underlying physics, and statistical noise due to finite-time averaging;
    \item we cannot trust the sensitivities of long-time averages or statistical averages to small parameter changes computed by using shadowing-based methods for sensitivity computation;
	\item it is possible for two models separated by tiny changes in parameter values to have drastically different statistical behavior. This means that the Lorenzian butterfly (a small perturbation) can significantly alter the climate of Texas (statistical/long-term average), and not just its weather (short-term events).
\end{itemize}
The paper is organized as follows. In section \ref{sec:shadowingIntro}, we introduce the dynamical systems concept of shadowing and its applications in the context of chaotic numerical simulations. In section \ref{sec:nonphysical}, we define the notion of physicality and give examples to illustrate the different ways in which to obtain nonphysical solutions. Section \ref{sec:shadowingnonphysical} is dedicated to several examples of perturbed tent maps that have nonphysical shadowing solutions. In the discussion in section \ref{sec:discussion}, we illustrate with an example that a small parameter perturbation can change the statistics significantly. The supplementary material, section \ref{sec:supp}, contains the numerical procedure to generate the probability distributions of the shadowing solutions.
